Palmer amaranth (\emph{Amaranthus palmeri}) adaptation to US Midwest agroecosystems

### 2021-08-23 Frontiers/Frontiers.tex
s

Abstract length and content varies depending on article type. Refer to 
\url{http://www.frontiersin.org/about/AuthorGuidelines} for abstract requirement
and length according to article type.

%All article types: you may provide up to 8 keywords; at least 5 are mandatory.
\tiny
 \keyFont{ \section{Keywords:} Evolution Flowering Management Pigweed Weed}

@@

Palmer amaranth is a fast growing summer annual forb indigenous to the
Sonoran Desert (Sauer, 1957). The species would eventually emerge as a
threat to US agriculture in the 1990s. Palmer amaranth weediness is
likely a result of human-assisted selection in combination with plant
biology. Farm mechanization, adoption of conservation agriculture (e.g.,
no-till), and reliance on herbicides for weed management are the main
human-mediated selection of Palmer amaranth into cropping systems (Ward
et al., 2013). On the other hand, Palmer amaranth is a prolific seed
producer with a C4 photosynthetic apparatus (Wang et al., 1992). With a
dioecy nature, Palmer amaranth male and female plants are obligate
outcrosser species, increasing the chances of exchanging adaptive traits
among plants (Jhala et al., 2021). Also, Palmer amaranth small seed size
(e.g, 1 mm) tend to thrive in no-tillage systems (Price et al., 2011),
and spread across locations through farm equipment (Sauer, 1972), manure
(Hartzler and Anderson, 2016), animals (Farmer et al., 2017). The
dispersal capacity of Palmer amaranth makes the species one of the most
successful cases of weed adaptation to current cropping systems.

Light and temperature are the main environment requirements for Palmer
amaranth successful grow and development (Jha et al., 2010). Palmer
amaranth is reported with an extended germination period (Ward et al.,
2013). Germination of Palmer amaranth was triggered by 18 C soil
temperature at 5 cm depth (Keeley et al., 1987), and optimal germination
and biomass production occurred at 35/30 C day and night temperatures
(Guo and Al-Khatib, 2003). In addition, Palmer amaranth establishment is
human-mediated by tillage timings and preemergence-applied herbicides
(Chahal et al., 2021), which can result in weed germination shifts
(Sbatella and Wilson, 2010). Water has not shown to limit Palmer
amaranth fitness. Under continuous water stress, Palmer amaranth
survived and produced at least 14000 seeds plant-1 (Chahal et al.,
2018). Seeds from Palmer amaranth growing with limited water conditions
were heavier, less dormant, and prompt for germination (Matzrafi et al.,
2021). Growing conditions and management practices also influence Palmer
amaranth sex dimorphism and flowering pattern (Korres et al., 2017;
Rumpa et al., 2019). Therefore, Palmer amaranth has shown plasticity to
evolve and fast adapt under selection pressure. Palmer amaranth
invasion/adaptation into new habitats is likely to increase with the
global temperature warming. Currently, it is estimated that the greatest
climatic risk of Palmer amaranth establishment are agronomic crops in
Australia and Sub-Sahara Africa (Kistner and Hatfield, 2018).
Temperature is a key factor limiting Palmer amaranth expansion to cooler
geographies (Briscoe Runquist et al., 2019); however, under future
climate change Palmer amaranth is likely to expand northward into Canada
and Northern Europe (Kistner and Hatfield, 2018; Briscoe Runquist et
al., 2019).

Palmer amaranth is already found in agronomic crops of South America
(Larran et al., 2017; Küpper et al., 2017) and Southern Europe (Milani
et al., 2021). In the United States, Palmer amaranth is well established
in the Cotton Belt (Garetson et al., 2019; Bagavathiannan and
Norsworthy, 2016) in the southern United States but its range is
expanding northward. For example, herbicide resistant Palmer amaranth is
widespread in Nebraska (Oliveira et al., 2021). There are some reported
cases of Palmer amaranth in Michigan (Kohrt et al., 2017) and
Connecticut (Aulakh et al., 2021). Also, it is estimated that Palmer
amaranth can cause high damage to soybean fields in Illinois (Davis et
al., 2015), which is concerning as soybean along with corn make most of
US Midwest agronomic hectares. In Iowa, a study showed that Palmer
amaranth is still not well adapted compared to waterhemp
(\emph{Amaranthus tuberculatus}) (Baker, 2021). Invasion and successful
eradication of Palmer amaranth is documented in Minnesota (Yu et al.,
2021). Palmer amaranth infestations have not been detected in Canada;
however, Palmer amaranth seeds were detected in sweet potato slips in
the country (Page et al., 2021). Palmer amaranth is still not as well
adapted and established to Northern as it is in the Southern North
America. Therefore, its range of expansion into new habitats is
increasing. It seems fated the need to manage Palmer amaranth in
agronomic crops throughout multiple environments in the near future.
Strategies on Palmer amaranth management should encompass the
agroecosystem level but not only attempts to eradicate the weed. Most
tactics to manage Palmer amaranth are based on technology fixes (Scott,
2011), which are short-term (e.g., herbicide and/or tillage) rather than
long-term weed management. Palmer amaranth management should be built on
minimizing the species ability to adapt, grow and develop into
agroecossystems.

In the southeastern US, early growing Palmer amaranth is well known to
have a higher impact on cotton yields compared to late established
plants (MacRae et al., 2013). In the northern states, Palmer amaranth
impact on the agroecosystem is recent. Studies investigating Palmer
amaranth in those locations are limited due to the plant classification
as noxious weed species in some northern states (Yu et al., 2021).
Nonetheless, the continuous Palmer amaranth dispersal and potential
establishment across the northern United States is concerning and
warrants investigations on species morphology in such environments.
Understanding Palmer amaranth biology and growing strategies under
different agroecosystems can enhance our knowledge on species adaptation
and management practices. It can also aid in designing proactive and
ecological tactics to limit the species range expansion, reduce its
negative impact, and developing resilient and sustainable farming
systems (MacLaren et al., 2020). Therefore, the objective of this study
was to investigate the flowering pattern, gender, biomass production,
and height of Palmer amaranth cohorts growing under corn, soybean and
bareground across five locations in the United States Midwest.

@@

A Palmer amaranth accession (Per1) from Perkins County, Nebraska. with
no reported herbicide resistance according to Oliveira et al. (2021) was
selected for this study. Three weeks prior to the onset of the field
experiments, seeds were planted in plastic trays containing potting-mix.
Emerged seedlings (1 cm) were transplanted into 200 cm-3 plastic pots (a
plant pot-1). Palmer amaranth seedlings were supplied with adequate
water and kept under greenhouse conditions at the University of
Wisconsin-Madison, University of Nebraska-Lincoln, and Western Illinois
University; and kept outdoors in the Perkins extension office at Grant,
NE until the 2-3 leaf stage (5 to 8 cm height) when they were
transported to the field.

@@

\multirow{-4}{*}{\raggedright\arraybackslash } & Organic matter (\%) &  & 3 &  &  & \\
\bottomrule
\end{tabular}}
\end{table}

@@

Palmer amaranth sex was fitted to a binary logistic regression
(Bangdiwala, 2018). Binary logistic regression is used for predicting
binary classes, such as the probability of a plant being female in a
dioecious species. Prior to the analysis, all missing values were
removed from the dataset. Also, data from Grant was not used in this
analysis due to the uniform plant harvesting at that location. The
complete dataset was splitted into 80\% train and 20\% test data. The
80\% train is used for the model training and the 20\% test is used for
checking how the model generalized on unseen dataset. With 80\% dataset,
a binary response variable, male (0) and female (1), was fitted to a
generalized linear model (\emph{glm} function) including day of year
harvest, height, weight, crop and month as independent variables
(without interaction). The model family was binomial with a logit
function. The model fit was assessed through pseudo R-squared values
(McFadden, Cox and Snell, Cragg and Uhler) and likelihood ratio using
\emph{nagelkerke} function (``rcompanion'' package). The marginal
effects computation was performed with Average Marginal Effects (AMEs)
at every observed value of x and average across the results (Leeper,
2017) using \emph{margins} function from ``margins'' package. The rest
20\% dataset was predicted using \emph{predict} function with a cutoff
estimation for male or female using \emph{performance} function. The
model quality prediction from the classification algorithm was measured
with precision (\emph{precision} function), recall (\emph{recall}
function) and F1-score (\emph{f\_meas} function) using the ``yardstick''
package. The precision determines the accuracy of positive predictions
(female plants), recall determines the fraction of positives that were
correctly identified, and F1-score is a weighted harmonic mean of
precision and recall with the best score of 1 and the worst score of 0.
F1-score conveys the balance between the precision and the recall
(Yacouby and Axman, 2020).

@@

\hypertarget{palmer-amaranth-gender}{%
\subsection*{Palmer amaranth gender}\label{palmer-amaranth-gender}}
\addcontentsline{toc}{subsection}{Palmer amaranth gender}

@@

\caption{\label{tab:unnamed-chunk-3}Average marginal means of gender logistic model. Factor pararemter values (e.g. crop) is shown related to soybean.}
\centering
\fontsize{10}{12}\selectfont
\begin{tabular}[t]{lrrrrrr}
\toprule
Term & AME & SE & Lower & Upper & Z-score & P-value\\
\midrule
crop\_bareground & -0.048 & 0.054 & -0.154 & 0.059 & -0.876 & 0.381\\

@@

Our study showed that Palmer amaranth biomass, height, flowering pattern
and gender varied within agroecossystems and cohort timings. In general,
first cohort of Palmer amaranth plants were heavier and taller when
compared to the second cohort. At first cohort, resources (e.g., soil
nutrients) and conditions (e.g., light) were more timely available for
the species. High biomass and taller Palmer amaranth plants are likely a
weed strategy to compete for light in between crop rows in absence of
canopy. In such conditions, Palmer amaranth showed an extraordinary
plasticity to adapt upon the agroecosystem. This is evident when
comparing Palmer amaranth canopy shape, and its extended flowering
pattern when growing into corn compared to soybean. The Palmer amaranth
competition strategy was to mimic the crop grow and development (Figure
6). These results suggests that Palmer amaranth can fast evolve
life-history traits to adapt into cropping systems and cultural
practices, which was also showed in a study varying nitrogen
fertilization (Bravo et al., 2018). Our results highlight the Palmer
amaranth as a threat to field crops as breeding more competitive crop
varieties is likely to select more competitive biotypes (Bravo et al.,
2017).

@@

Seed production was not evaluated due to plant harvest at initiation of
flowering. Nonetheless, it is well documented a strong positive
correlation between Palmer amaranth biomass and seed production
(Schwartz et al., 2016; Spaunhorst et al., 2018). I our study, Palmer
amaranth growing at first cohort accumulated an overall 36\% more
biomass when compared to second cohort. Therefore, Palmer amaranth
plants growing in the second cohort is likely to produce less seeds
regardless the cropping system. Our observation is consistent with the
findings that first Palmer amaranth cohort produced 50\% more seeds per
plant than Palmer amaranth plants established six weeks later in
bareground (Webster and Grey, 2015). Still, seed production at second
cohort is likely to replenish the soil seedbank. Seed production and
deposition in the seedbank is also a key factor for species perpetuation
(Menges, 1987). Palmer amaranth can produce hundred thousands seeds per
plant (Schwartz et al., 2016; Keeley et al., 1987), and stay viable
buried in the seedbank for at least 36 months (Sosnoskie et al., 2013).
Therefore, preventing Palmer amaranth seed production or/and seed
migration to its non-native habitat is an essential strategy to minimize
weed impact into agroecossystem (Davis et al., 2015).

An ecological approach to reduce seed production in Palmer amaranth is
understanding its flowering window. Our study suggests that Palmer
amaranth flowering was slightly influenced by cropping systems and
cohort timings. Palmer amaranth growing in bareground and corn resulted
in the overall shortest and longest flowering window, respectively. When
growing in soybean, Palmer amaranth flowering window was similar to
bareground at first cohort but similar to corn at second cohort timing.
Plant flowering initiation is complex and depends on biological and
ecological factors (Lang, 1965). We hypothesize that when growing in
high competition (e.g., second cohort), Palmer amaranth plants tend to
initiate flowering early, as well as having an extended flowering
window. Early flower initiation is plant strategy when growing in stress
conditions. For example, when growing under water stress, early
flowering in Palmer amaranth resulted in a mismatch between female and
male plants by seven days (Mesgaran et al., 2021). A mismatch in Palmer
amaranth male and female flowering period can minimize plant outcross,
and thus reduce plant seed production and exchange of resistant alleles
(Jhala et al., 2021). Sex dimorphism manipulation is considered a
potential ecological pest control (McFarlane et al., 2018; Schliekelman
et al., 2005).

The mechanisms of sex-determination in plant species is intriguing and
arouse the curiosity of many scientists, including Charles Darwin
(Darwin, 1888). In our study, the gender model performance was decent
(AIC 0.64) considering the biology of plant flowering. A 1:1 male and
female sex ratio is a general evolutionary stable strategy for plant
species perpetuation (Fisher, 1930). However, a slight deviation from
1:1 sex ratio might occur in some dioecious species. For example, the
dioecious \emph{Halophila stipulacea} is a female-biased plant in its
native habitat, but the naturalized \emph{H. stipulacea} have a 1:1
ratio (Nguyen et al., 2018). Naturalized of \emph{H. stipulacea} reduced
female-male ratio to expand into its non-native habitat (Nguyen et al.,
2018). Also, biotic and/or abiotic stress can influence plant sex
determination. Palmer amaranth male-to-female ratio was greater under
high plant densities (Korres and Norsworthy, 2017) and after herbicide
application (Rumpa et al., 2019). Our model estimated that late
flowering, heavier and taller Palmer amaranth plants deviated from 1:1
ratio in favor to female plants. It was reported that female Palmer
amaranth plants invested more in height, stem and biomass while male
invested more in leaf area and leaf dry weight under nutrient deficiency
(Korres et al., 2017). Our model also estimated more female plants in
soybean and bareground compared to corn, which might linked to plant
competition strategy in each agroecossystem. Our results showed the
influence of life-history and ecological traits on sexual dimorphism in
Palmer amaranth. Sexual dimorphism is documented in other dioecious
species (Barrett and Hough, 2013). For example, stronger female plant
competition and greater male tolerance to herbivory was reported in
\emph{Spinacia oleracea} (Pérez-Llorca and Sánchez Vilas, 2019).
Research on candidate genes for sex determination in \emph{Amaranthus}
species are currently underway but it is far to complete (Montgomery et
al., 2021, 2019). Further studies are also needed to understand the
ecological basis of Palmer amaranth flowering, including the plant
behavior under climate change.

Our study demonstrated the short-term Palmer amaranth plasticity to grow
and develop into cropping-systems. Is likely that Palmer amaranth range
will continue to expanding to new geographies. Therefore, preventive
management is a priority to minimizing Palmer amaranth dispersal.
Reactive management should focus on early-season management programs,
which would have a large negative effect on Palmer amaranth growth and
development. Tactics that promote early-season crop advantage against
Palmer amaranth, including early crop planting, crop rotation (Oliveira
et al., 2021), plant width, preemergence applied herbicide (Sanctis et
al., 2021), and crop residue (e.g.~cover crops) would minimize the
negative impact of Palmer amaranth in agroecossystems. The
aggressiveness and differential Palmer amaranth adaptation to
agroecosystem is striking and require national efforts to minimize the
species impact on economy and sustainability.

@@

\hypertarget{supplemental-data}{%
\section*{Supplemental Data}\label{supplemental-data}}
\addcontentsline{toc}{section}{Supplemental Data}

Supplementary Material should be uploaded separately on submission, if
there are Supplementary Figures, please include the caption in the same
file as the figure. LaTeX Supplementary Material templates can be found
in the Frontiers LaTeX folder

@@

@@

\leavevmode\hypertarget{ref-sanctis2021}{}%
Sanctis, J. H. S. de, Barnes, E. R., Knezevic, S. Z., Kumar, V., and
Jhala, A. J. (2021). Residual herbicides affect critical time of
{Palmer} amaranth removal in soybean. \emph{Agronomy Journal} 113,
1920--1933.
doi:\href{https://doi.org/10.1002/agj2.20615}{10.1002/agj2.20615}.

@@

\caption{Palmer amaranth biomass (A) and height (B) growing in corn, bareground, and soybean across Arlington, WI, Clay Center, NE, Grant, NE, Lincoln, NE and Macomb, IL}\label{fig:Figure-3}
\end{figure}

>>>>

Palmer amaranth (\emph{Amaranthus palmeri} S. Watson) is one of the most troublesome weed species in the United States. Palmer amaranth is endemic in the Southern United States but its range is expanding northward. Palmer amaranth dispersal warrants studies assessing species adaptation into new geographies. A study was conducted to investigate morphology, flowering and sex dimorphism from cohorts of Palmer amaranth growing under corn, soybean, and bareground across five locations of United States Midwest. In general, results demonstrated that first cohort of Palmer amaranth, established in June, produced 42\% more biomass than plants from second cohort (established in July). At first cohort, Palmer amaranth produced 75.5 g plant\textsuperscript{-1} in bareground, 28.3 g plant\textsuperscript{-1} in soybean and 16.3 g plant\textsuperscript{-1} in corn, whereas Palmer amaranth produced 62.6, 6.3, and 1.4 g plant\textsuperscript{-1} at second cohort in bareground, soybean and corn, respectively. The plant height was more impacted by Palmer amaranth growing in corn. Palmer amaranth plants averaged 85.2 cm tall in the first cohort but 38.2 cm tall in the second cohort in corn. Moreover, Palmer amaranth flowering window shifted according to crop and cohort timings. Palmer amaranth growing in intense competition, such as under low light in corn, resulted in the longest flowering window. Also, Palmer amaranth sex dimorphism was slightly influenced by day of year, weight and height. The model estimated that probability of being a female plant increase as biomass and height increases. Our results showed the fast adaptation and plasticity of Palmer amaranth to grow and adapt to cropping systems. Palmer amaranth is likely to continue its expansion northward. Therefore, preventing plant dispersal into new habitat is the most effective management strategy. Reactive management to reduce Palmer amaranth impact on cropping systems should encompass diversity of tactics that minimize the species ability establish into cropping systems, including crop rotation (not only corn and soybean), early/late crop planting, row spacing, and cover crops. Our results also shed a light on sex dimorphism as a future Palmer amaranth ecological management.

\tiny
 \keyFont{ \section{Keywords:} Evolution Flowering Management Pigweed Weed}

@@

Palmer amaranth is a fast growing summer annual forb indigenous to the
Sonoran Desert (Sauer, 1957). The species would eventually emerge as a
threat to US agriculture in the 1990s. Palmer amaranth weediness is
likely a result of human-assisted selection in combination with plant
biology. Farm mechanization, adoption of conservation agriculture (e.g.,
no-till), and reliance on herbicides for weed management are the main
human-mediated selection of Palmer amaranth into cropping systems (Ward
et al., 2013). On the other hand, Palmer amaranth is a prolific seed
producer with a C4 photosynthetic apparatus (Wang et al., 1992). With a
dioecy nature, Palmer amaranth male and female plants are obligate
outcrosser species, increasing the chances of exchanging adaptive traits
among plants (Jhala et al., 2021; Oliveira et al., 2018). Also, Palmer
amaranth small seed size (e.g, 1 mm) tend to thrive in no-tillage
systems (Price et al., 2011), and spread across locations through farm
equipment (Sauer, 1972), manure (Hartzler and Anderson, 2016), animals
(Farmer et al., 2017). The dispersal capacity of Palmer amaranth makes
the species one of the most successful cases of weed adaptation to
current cropping systems.

Light and temperature are the main environment requirements for Palmer
amaranth successful grow and development (Jha et al., 2010). Palmer
amaranth is reported with an extended germination period (Ward et al.,
2013). Germination of Palmer amaranth was triggered by 18 C soil
temperature at 5 cm depth (Keeley et al., 1987), and optimal germination
and biomass production occurred at 35/30 C day and night temperatures
(Guo and Al-Khatib, 2003). In addition, Palmer amaranth establishment is
human-mediated by tillage timings and preemergence-applied herbicides
(Chahal et al., 2021), which can result in weed germination shifts
(Sbatella and Wilson, 2010). Water has not shown to limit Palmer
amaranth fitness. Under continuous water stress, Palmer amaranth
survived and produced at least 14000 seeds plant\textsuperscript{-1}
(Chahal et al., 2018). Seeds from Palmer amaranth growing with limited
water conditions were heavier, less dormant, and prompt for germination
(Matzrafi et al., 2021). Growing conditions and management practices
also influence Palmer amaranth sex dimorphism and flowering pattern
(Korres et al., 2017; Rumpa et al., 2019). Therefore, Palmer amaranth
has shown plasticity to evolve and fast adapt under the current
conditions. Further scenarios show that global temperature warming can
impact agriculture, and promote niches for Palmer amaranth
invasion/adaptation into new environments. Currently, it is estimated
that the greatest climatic risk of Palmer amaranth establishment are
agronomic crops in Australia and Sub-Sahara Africa (Kistner and
Hatfield, 2018). Temperature is a key factor limiting Palmer amaranth
expansion to cooler geographies (Briscoe Runquist et al., 2019);
however, under future climate change Palmer amaranth is likely to expand
northward into Canada and Northern Europe (Kistner and Hatfield, 2018;
Briscoe Runquist et al., 2019).

Palmer amaranth is already found in agronomic crops of South America
(Larran et al., 2017; Küpper et al., 2017) and Southern Europe (Milani
et al., 2021). In the United States, Palmer amaranth is well established
in the Cotton Belt (Garetson et al., 2019; Bagavathiannan and
Norsworthy, 2016) in the southern United States but its range is
expanding northward. For example, herbicide resistant Palmer amaranth is
widespread in Nebraska (Oliveira et al., 2021). There are some reported
cases of Palmer amaranth in Michigan (Kohrt et al., 2017) and
Connecticut (Aulakh et al., 2021). Also, it is estimated that Palmer
amaranth can cause high damage to soybean fields in Illinois (Davis et
al., 2015), which is concerning as soybean along with corn make most of
US Midwest agronomic hectares. In Iowa, a study showed that Palmer
amaranth is still not well adapted compared to waterhemp
(\emph{Amaranthus tuberculatus}) (Baker, 2021). Invasion and successful
eradication of Palmer amaranth is documented in Minnesota (Yu et al.,
2021). Palmer amaranth infestations have not been detected in Canada;
however, Palmer amaranth seeds were detected in sweet potato slips in
the country (Page et al., 2021). Palmer amaranth is still not as well
adapted and established to Northern as it is in the Southern North
America. Therefore, Palmer amaranth range of expansion into new habitats
is increasing. It seems fated the need to manage Palmer amaranth in
agronomic crops throughout multiple environments in the near future.
Strategies on Palmer amaranth management should encompass the
agroecosystem level but not only attempts to eradicate the weed. Most
tactics to manage Palmer amaranth are based on technology fixes (Scott,
2011), which are short-term (e.g., herbicide and/or tillage) rather than
long-term weed management. Palmer amaranth management should be built on
minimizing the species ability to adapt, grow and develop into
agroecossystems.

In the southeastern US, early growing Palmer amaranth is well known to
have a higher impact on cotton yields compared to late established
plants (MacRae et al., 2013). In the northern states, Palmer amaranth
impact on the agroecosystem is recent. Studies investigating Palmer
amaranth in those locations are limited due to the plant classification
as noxious weed species in some northern states (Yu et al., 2021).
Nonetheless, the continuous Palmer amaranth dispersal and potential
establishment across the northern United States is concerning and
warrants investigations on species morphology in such environments.
Understanding Palmer amaranth biology and growing strategies under
different agroecosystems can enhance our knowledge on species adaptation
and management practices. It can also aid in designing proactive and
ecological tactics to limit the species range expansion, reduce its
negative impact, and developing resilient and sustainable farming
systems (MacLaren et al., 2020). Therefore, the objective of this study
was to investigate the flowering pattern, sex dimorphism, biomass
production, and height of Palmer amaranth cohorts growing under corn,
soybean and bareground across five locations in the United States
Midwest.

@@

A Palmer amaranth accession (Per1) from Perkins County, Nebraska. with
no reported herbicide resistance according to Oliveira et al. (2021) was
selected for this study. Three weeks prior to the onset of the field
experiments, seeds were planted in plastic trays containing potting-mix.
Emerged seedlings (1 cm) were transplanted into 200
cm\textsuperscript{-3} plastic pots (a plant pot\textsuperscript{-1}).
Palmer amaranth seedlings were supplied with adequate water and kept
under greenhouse conditions at the University of Wisconsin-Madison,
University of Nebraska-Lincoln, and Western Illinois University; and
kept outdoors at the Perkins extension office in Grant, NE until the 2-3
leaf stage (5 to 8 cm height) when they were transported to the field.

@@

\multirow{-4}{*}{\raggedright\arraybackslash } & Organic matter (\%) &  & 3 &  &  & \\
\bottomrule
\multicolumn{7}{l}{\rule{0pt}{1em}\textsuperscript{a} glyphoste, 840 g ae ha \textsuperscript{b} S-metolachor, g ai ha \textsuperscript{c} S-metolachlor + trazine + mesotrione, + bicyclopyrone, g ai ha \textsuperscript{d} saflufenacil + imazethapyr + pyroxasulfone, g ai ha}\\
\end{tabular}}
\end{table}

@@

Palmer amaranth sex dimorphism was fitted to a binary logistic
regression (Bangdiwala, 2018). Binary logistic regression is used for
predicting binary classes, such as the probability of a plant being
female in a dioecious species. Prior to the analysis, all missing values
were removed from the dataset. Also, data from Grant was not used in
this analysis due to the uniform plant harvesting at that location. The
complete dataset was splitted into 80\% train and 20\% test data. The
80\% train is used for the model training and the 20\% test is used for
checking how the model generalized on unseen dataset. With 80\% dataset,
a binary response variable, male (0) and female (1), was fitted to a
generalized linear model (\emph{glm} function) including day of year
harvest, height, weight, crop and month as independent variables
(without interaction). The model family was binomial with a logit
function. The model fit was assessed through pseudo R-squared values
(McFadden, Cox and Snell, Cragg and Uhler) and likelihood ratio using
\emph{nagelkerke} function (``rcompanion'' package). The marginal
effects computation was performed with Average Marginal Effects (AMEs)
at every observed value of x and average across the results (Leeper,
2017) using \emph{margins} function from ``margins'' package. The rest
20\% dataset was predicted using \emph{predict} function with a cutoff
estimation for male or female using \emph{performance} function. The
model quality prediction from the classification algorithm was measured
with precision (\emph{precision} function), recall (\emph{recall}
function) and F1-score (\emph{f\_meas} function) using the ``yardstick''
package. The precision determines the accuracy of positive predictions
(female plants), recall determines the fraction of positives that were
correctly identified, and F1-score is a weighted harmonic mean of
precision and recall with the best score of 1 and the worst score of 0.
F1-score conveys the balance between the precision and the recall
(Yacouby and Axman, 2020).

@@

\hypertarget{palmer-amaranth-sex-dimorphism}{%
\subsection*{Palmer amaranth sex
dimorphism}\label{palmer-amaranth-sex-dimorphism}}
\addcontentsline{toc}{subsection}{Palmer amaranth sex dimorphism}

@@

\caption{\label{tab:unnamed-chunk-3}Average marginal means of sex dimorphism logistic model. Factor pararemter values (e.g. crop) is shown related to soybean.}
\centering
\fontsize{10}{12}\selectfont
\begin{tabular}[t]{lrrrrrr}
\toprule
Term & AME & SE & Lower & Upper & Z-score & P-value\\
\midrule
crop\_bareground & -0.048 & 0.054 & -0.154 & 0.059 & -0.876 & 0.381\\

@@

Our study showed that Palmer amaranth biomass, height, flowering pattern
and sex varied within crops and cohort timings. In general, first cohort
of Palmer amaranth plants were heavier and taller when compared to the
second cohort. At first cohort, resources (e.g., soil nutrients) and
conditions (e.g., light) were more timely available for the species.
High biomass and taller Palmer amaranth plants are likely a weed
strategy to compete for light in between crop rows in absence of canopy.
In such conditions, Palmer amaranth showed an extraordinary plasticity
to adapt upon the agroecosystem. This is evident when comparing Palmer
amaranth canopy shape, and its extended flowering pattern when growing
into corn compared to soybean. The Palmer amaranth competition (e.g.,
light) strategy was to mimic the crop grow and development (Figure 6).
These results suggests that Palmer amaranth can fast evolve life-history
traits to adapt into cropping systems and cultural practices, which was
also showed in a study varying nitrogen fertilization (Bravo et al.,
2018). Our results highlight the Palmer amaranth as a threat to field
crops as breeding more competitive crop varieties is likely to select
more competitive biotypes (Bravo et al., 2017).

@@

Seed production was not evaluated due to plant harvest at initiation of
flowering. Nonetheless, it is well documented a strong positive
correlation between Palmer amaranth biomass and seed production
(Schwartz et al., 2016; Spaunhorst et al., 2018). I our study, Palmer
amaranth growing at first cohort accumulated an overall 42\% more
biomass when compared to second cohort. Therefore, Palmer amaranth
plants growing in the second cohort is likely to produce less seeds
regardless the cropping system. Our observation is consistent with the
findings that first Palmer amaranth cohort produced 50\% more seeds per
plant than Palmer amaranth plants established six weeks later in
bareground (Webster and Grey, 2015). Still, seed production at second
cohort is likely to replenish the soil seedbank. Seed production and
deposition in the seedbank is also a key factor for species perpetuation
(Menges, 1987). Palmer amaranth can produce hundred thousands seeds per
plant (Schwartz et al., 2016; Keeley et al., 1987), and stay viable
buried in the seedbank for at least 36 months (Sosnoskie et al., 2013).
Therefore, preventing Palmer amaranth seed production or/and seed
migration to its non-native habitat is an essential strategy to minimize
weed impact into agroecossystem (Davis et al., 2015).

An ecological approach to reduce seed production in Palmer amaranth is
understanding its flowering window. Our study suggests that Palmer
amaranth flowering was slightly influenced by cropping systems and
cohort timings. Palmer amaranth growing in bareground and corn resulted
in the overall shortest and longest flowering window, respectively. When
growing in soybean, Palmer amaranth flowering window was similar to
bareground at first cohort but similar to corn at second cohort timing.
Plant flowering initiation is complex and depends on biological and
ecological factors (Lang, 1965). We hypothesize that when growing in
high competition (e.g., second cohort) and/or under low light/short day
length, Palmer amaranth plants tend to initiate flowering early, as well
as having an extended flowering window. Early flower initiation is plant
strategy when growing in stress conditions. For example, when growing
under water stress, early flowering in Palmer amaranth resulted in a
mismatch between female and male plants by seven days (Mesgaran et al.,
2021). A mismatch in Palmer amaranth male and female flowering period
can minimize plant outcross, and thus reduce plant seed production and
exchange of resistant alleles (Jhala et al., 2021). Sex dimorphism
manipulation is considered a potential ecological pest control
(McFarlane et al., 2018; Schliekelman et al., 2005).

The mechanisms of sex-determination in plant species is intriguing and
arouse the curiosity of many scientists, including Charles Darwin
(Darwin, 1888). In our study, the sex dimorphism model performance was
decent (AIC 0.64) considering the biology of plant flowering. A 1:1 male
and female sex ratio is a general evolutionary stable strategy for plant
species perpetuation (Fisher, 1930). However, a slight deviation from
1:1 sex ratio might occur in some dioecious species. For example, the
dioecious \emph{Halophila stipulacea} is a female-biased plant in its
native habitat, but the naturalized \emph{H. stipulacea} have a 1:1
ratio (Nguyen et al., 2018). Naturalized of \emph{H. stipulacea} reduced
female-male ratio to expand into its non-native habitat (Nguyen et al.,
2018). Also, biotic and/or abiotic stress can influence plant sex
determination. Palmer amaranth male-to-female ratio was greater under
high plant densities (Korres and Norsworthy, 2017) and after herbicide
application (Rumpa et al., 2019). Our model estimated that late
flowering, heavier and taller Palmer amaranth plants deviated from 1:1
ratio in favor to female plants. It was reported that female Palmer
amaranth plants invested more in height, stem and biomass while male
invested more in leaf area and leaf dry weight under nutrient deficiency
(Korres et al., 2017). Our model also estimated more female plants in
soybean and bareground compared to corn, which might linked to plant
competition strategy in each agroecossystem. Our results showed the
influence of life-history and ecological traits on sexual dimorphism in
Palmer amaranth. Sexual dimorphism is documented in other dioecious
species (Barrett and Hough, 2013). For example, stronger female plant
competition and greater male tolerance to herbivory was reported in
\emph{Spinacia oleracea} (Pérez-Llorca and Sánchez Vilas, 2019).
Research on candidate genes for sex determination in \emph{Amaranthus}
species are currently underway but it is far to complete (Montgomery et
al., 2021, 2019). Further studies are also needed to understand the
ecological basis of Palmer amaranth flowering, including the plant
behavior under climate change.

Our study demonstrated the Palmer amaranth plasticity to grow and
develop into cropping-systems. Is likely that Palmer amaranth range will
continue to expanding into new geographies. Therefore, preventive
management is a priority to minimizing Palmer amaranth dispersal.
Reactive management should focus on early-season management programs,
which would have a large negative effect on Palmer amaranth growth and
development. Tactics that promote early-season crop advantage against
Palmer amaranth, including early crop planting, crop rotation, plant
width, preemergence applied herbicide, and crop residue (e.g.~cover
crops) would minimize the negative impact of Palmer amaranth in
agroecossystems. The aggressiveness and differential Palmer amaranth
adaptation to agroecosystem is striking and require national efforts to
minimize the species impact on economy and sustainability.

@@

@@

\leavevmode\hypertarget{ref-oliveira2018}{}%
Oliveira, M. C., Gaines, T. A., Patterson, E. L., Jhala, A. J., Irmak,
S., Amundsen, K., and Knezevic, S. Z. (2018). Interspecific and
intraspecific transference of metabolism-based mesotrione resistance in
dioecious weedy {Amaranthus}. \emph{The Plant Journal} 96, 1051--1063.
doi:\href{https://doi.org/10.1111/tpj.14089}{10.1111/tpj.14089}.

@@

@@

\caption{Palmer amaranth biomass (A) and height (B) growing in corn, bareground, and soybean nested across Arlington, WI, Clay Center, NE, Grant, NE, Lincoln, NE and Macomb, IL}\label{fig:Figure-3}
\end{figure}
